High-Dimensional Regularized Discriminant Analysis

### 2016-11-12 reports/paper-hdrda.Rnw

\author[jar]{John A. Ramey\cortext[cor1]{Corresponding author}}
\author[cks]{Caleb K. Stein}
\author[pdy]{Phil D. Young}
\author[dmy]{Dean M. Young}
\address[jar]{uStudio, Inc.}
\address[cks]{Myeloma Institute, University of Arkansas for Medical Sciences}
\address[pdy]{Department of Management and Information Systems, Baylor University}
\address[dmy]{Department of Statistical Science, Baylor University}

@@

As we have discussed above, several eigenvalue adjustment methods have been
proposed that increase eigenvalues (approximately) equal to 0. To further
improve the estimation of $\bm \Sigma_k$ and to stabilize the estimator's
inverse, we define the eigenvalue adjustment of
\eqref{eq:sig-lambda-alternative} as
\begin{linenomath*}
\begin{align}
	\tilde{\bm \Sigma}_k \coloneqq \alpha_k \widehat{\bm \Sigma}_k(\lambda) + \gamma \bm I_p,\label{eq:hdrda-cov}
\end{align}
\end{linenomath*}
where $\alpha_k \ge 0$ and $\gamma \ge 0$ is an eigenvalue-shrinkage
constant. Thus, the \emph{pooling} parameter $\lambda$ controls the amount of
estimation information borrowed from $\widehat{\bm \Sigma}$ to estimate $\bm
\Sigma_k$, and the \emph{shrinkage} parameter $\gamma$ determines the degree of
eigenvalue shrinkage. The choice of $\alpha_k$ allows for a flexible formulation
of covariance-matrix estimators. For instance, if $\alpha_k = 1$, $k = 1,
\ldots, K$, then \eqref{eq:hdrda-cov} resembles \eqref{eq:ridge-estimator}.
Similarly, if $\alpha_k = 1 - \gamma$, then \eqref{eq:hdrda-cov} has a form
comparable to the \emph{RDA} classifier from \cite{Friedman:1989tm}.
Substituting \eqref{eq:hdrda-cov} into \eqref{eq:qda}, we define the
\emph{HDRDA} classifier
\begin{linenomath*}
\begin{align}
	D_{HDRDA}(\bm x) = \argmin_{k}  (\bm x - \xbar_k)^{T}\tilde{\bm \Sigma}_k^{+}(\bm x - \xbar_k)  + \log |\tilde{\bm \Sigma}_k|. \label{eq:hdrda}
\end{align}
\end{linenomath*}
For $\gamma > 0$, $\tilde{\bm \Sigma}_k$ is nonsingular such that $\tilde{\bm
  \Sigma}_k^{-1}$ can be substituted for $\tilde{\bm \Sigma}_k^{+}$ in
\eqref{eq:hdrda}. If $\gamma = 0$, we explicitly set $|\tilde{\bm \Sigma}_k|$
equal to the product of the positive eigenvalues of $\tilde{\bm
  \Sigma}_k$. Following \cite{Friedman:1989tm}, we select $\lambda$ and $\gamma$
from a grid of candidate models via cross-validation. We provide an
implementation of \eqref{eq:hdrda} in the {\tt hdrda} function contained in the
{\tt sparsediscrim} R package, which is available on CRAN.

@@

@@

In this section, we compare our proposed classifier with four classifiers
recently proposed for small-sample, high-dimensional data along with the
random-forest classifier from \cite{Breiman:2001fb} using version 3.2.2 of the
open-source statistical software {\tt R}. Within our study, we included
penalized linear discriminant analysis from \cite{Witten:2011kc}, implemented in
the {\tt penalizedLDA} package. We also considered shrunken centroids
regularized discriminant analysis from \cite{Guo:2007te} in the {\tt rda}
package. Because the {\tt rda} package does not perform the authors' ``Min-Min''
rule automatically, we applied this rule within our {\tt R} code. We included
two modifications of diagonal linear discriminant analysis from
\cite{Tong:2012hw} and \cite{Pang:2009ik}, where the former employs an improved
mean estimator and the latter utilizes an improved variance estimator. Both
classifiers are available in the {\tt sparsediscrim} package. Finally, we
incorporated the random forest as a benchmark based on the findings of
\cite{FernandezDelgado:2014ul}, who concluded that the random forest is often
superior to other classifiers in benchmark studies. We used the implementation
of the random-forest classifier from the {\tt randomForest} package with 250
trees and 100 maximum nodes. For each classifer we explicitly set prior
probabilities as equal, if applicable. All other classifier options were set to
their default settings. Below, we refer to each classifier by the first author's
surname. All simulations were conducted on an Amazon Elastic Compute Cloud (EC2)
{\tt c4.4xlarge} instance. Our analyses can be reproduced using the code
available in the \url{https://github.com/ramhiser/paper-hdrda} repository.

For the \emph{HDRDA} classifier in \eqref{eq:hdrda-decomposed}, we examined the
classification performance of two models. For the first \emph{HDRDA} model, we
set $\alpha_k = 1$, $k = 1, \ldots, K$, so that the covariance-matrix estimator
\eqref{eq:hdrda-cov} resembled \eqref{eq:ridge-estimator}. We estimated
$\lambda$ from a grid of 21 equidistant candidate values between 0 and 1,
inclusively. Similarly, we estimated $\gamma$ from a grid consisting of the
values $10^{-1}, \ldots, 10^4$, and $10^5$. We selected optimal estimates of
$\lambda$ and $\gamma$ using $V$-fold cross-validation with $V = 10$
\citep{Hastie:2008dt}. For the second model, we set $\alpha_k = 1 - \gamma$, $k
= 1, \ldots, K$, to resemble \citeauthor{Friedman:1989tm}'s parameterization,
and we estimated both $\lambda$ and $\gamma$ from a grid of 21 equidistant
candidate values between 0 and 1, inclusively.

>>>>

\author[jar]{John A. Ramey\cortext[cor1]{Corresponding author}}
\author[cks]{Caleb K. Stein}
\author[pdy]{Phil D. Young}
\author[dmy]{Dean M. Young}
\address[jar]{uStudio Labs}
\address[cks]{Myeloma Institute, University of Arkansas for Medical Sciences}
\address[pdy]{Department of Management and Information Systems, Baylor University}
\address[dmy]{Department of Statistical Science, Baylor University}

@@

As we have discussed above, several eigenvalue adjustment methods have been
proposed that increase eigenvalues (approximately) equal to 0. To further
improve the estimation of $\bm \Sigma_k$ and to stabilize the estimator's
inverse, we define the eigenvalue adjustment of
\eqref{eq:sig-lambda-alternative} as
\begin{linenomath*}
\begin{align}
	\tilde{\bm \Sigma}_k \coloneqq \alpha_k \widehat{\bm \Sigma}_k(\lambda) + \gamma \bm I_p,\label{eq:hdrda-cov}
\end{align}
\end{linenomath*}
where $\alpha_k \ge 0$ and $\gamma \ge 0$ is an eigenvalue-shrinkage
constant. Thus, the \emph{pooling} parameter $\lambda$ controls the amount of
estimation information borrowed from $\widehat{\bm \Sigma}$ to estimate $\bm
\Sigma_k$, and the \emph{shrinkage} parameter $\gamma$ determines the degree of
eigenvalue shrinkage. The choice of $\alpha_k$ allows for a flexible formulation
of covariance-matrix estimators. For instance, if $\alpha_k = 1$, $k = 1,
\ldots, K$, then \eqref{eq:hdrda-cov} resembles \eqref{eq:ridge-estimator}.
Similarly, if $\alpha_k = 1 - \gamma$, then \eqref{eq:hdrda-cov} has a form
comparable to the \emph{RDA} classifier from \cite{Friedman:1989tm}.
Substituting \eqref{eq:hdrda-cov} into \eqref{eq:qda}, we define the
\emph{HDRDA} classifier
\begin{linenomath*}
\begin{align}
	D_{HDRDA}(\bm x) = \argmin_{k}  (\bm x - \xbar_k)^{T}\tilde{\bm \Sigma}_k^{+}(\bm x - \xbar_k)  + \log |\tilde{\bm \Sigma}_k|. \label{eq:hdrda}
\end{align}
\end{linenomath*}
For $\gamma > 0$, $\tilde{\bm \Sigma}_k$ is nonsingular such that $\tilde{\bm
  \Sigma}_k^{-1}$ can be substituted for $\tilde{\bm \Sigma}_k^{+}$ in
\eqref{eq:hdrda}. If $\gamma = 0$, we explicitly set $|\tilde{\bm \Sigma}_k|$
equal to the product of the positive eigenvalues of $\tilde{\bm
  \Sigma}_k$. Following \cite{Friedman:1989tm}, we select $\lambda$ and $\gamma$
from a grid of candidate models via cross-validation \citep{Hastie:2008dt}. We provide an
implementation of \eqref{eq:hdrda} in the {\tt hdrda} function contained in the
{\tt sparsediscrim} R package, which is available on CRAN.

@@

\section{Timing Comparisons between RDA and HDRDA}
\label{sec:timing-comparisons}

In this section, we demonstrate that the computational performance of the
training and model selection employed in the \emph{HDRDA} classifier is
substantially faster than that of the \emph{RDA} classifier on small-sample,
high-dimensional data sets. The relative difference in runtime between the two
classifiers drastically increases as the number of features increases. To
compare the two classifiers, we generate 25 observations from each of $K=4$
multivariate normal populations with mean vectors $\bm \mu_1 = -3 \cdot \bm
1_p$, $\bm \mu_2 = -\bm 1_p$, $\bm \mu_3 = \bm 1_p$, and $\bm \mu_4 = 3 \cdot
\bm 1_p$. We set the covariance matrix of each population to the $p \times p$
identity matrix. For each data set generated, we estimate the $\lambda$ and
$\gamma$ parameters for both classifiers using a grid of 21 equidistant
candidate values between 0 and 1, inclusively. We set $\alpha_k = 1 - \gamma$,
$k = 1, \ldots, K$, in the \emph{HDRDA} classifier to resemble
\citeauthor{Friedman:1989tm}'s \emph{RDA} parameterization. At each pair of
$\lambda$ and $\gamma$, we compute the $V$-fold cross-validation error rate with
$V = 10$ \citep{Hastie:2008dt}. The optimal model is the pair of $\lambda$ and
$\gamma$ that minimizes the $V$-fold cross-validation error rate.

Here, our goal is to compare the computational runtime of the model selection
for both classifiers. We compare the runtime of both classifiers by increasing
the number of features from $p = 500$ to $p = 5000$ in increments of 500. We
generate 100 data sets for each value of $p$ and compute the training and model
selection runtime of both classifiers. Our timing comparisons are based on our
\emph{HDRDA} implementation in the {\tt sparsediscrim} R package with the
standard \emph{RDA} implementation in the {\tt klaR} R package. All timing
comparisons were conducted on an Amazon Elastic Compute Cloud (EC2) {\tt
  c4.4xlarge} instance using version 3.3.1 of the open-source statistical
software {\tt R}. Our analyses can be reproduced using the code available in the
\url{https://github.com/ramhiser/paper-hdrda} repository.

\subsection{Timing Comparison Results}

TODO

@@

In this section, we compare our proposed classifier with four classifiers
recently proposed for small-sample, high-dimensional data along with the
random-forest classifier from \cite{Breiman:2001fb} using version 3.3.1 of the
open-source statistical software {\tt R}. Within our study, we included
penalized linear discriminant analysis from \cite{Witten:2011kc}, implemented in
the {\tt penalizedLDA} package. We also considered shrunken centroids
regularized discriminant analysis from \cite{Guo:2007te} in the {\tt rda}
package. Because the {\tt rda} package does not perform the authors' ``Min-Min''
rule automatically, we applied this rule within our {\tt R} code. We included
two modifications of diagonal linear discriminant analysis from
\cite{Tong:2012hw} and \cite{Pang:2009ik}, where the former employs an improved
mean estimator and the latter utilizes an improved variance estimator. Both
classifiers are available in the {\tt sparsediscrim} package. Finally, we
incorporated the random forest as a benchmark based on the findings of
\cite{FernandezDelgado:2014ul}, who concluded that the random forest is often
superior to other classifiers in benchmark studies. We used the implementation
of the random-forest classifier from the {\tt randomForest} package with 250
trees and 100 maximum nodes. For each classifer we explicitly set prior
probabilities as equal, if applicable. All other classifier options were set to
their default settings. Below, we refer to each classifier by the first author's
surname. All simulations were conducted on an Amazon EC2 {\tt c4.4xlarge}
instance. Our analyses can be reproduced using the code available in the
\url{https://github.com/ramhiser/paper-hdrda} repository.

For the \emph{HDRDA} classifier in \eqref{eq:hdrda-decomposed}, we examined the
classification performance of two models. For the first \emph{HDRDA} model, we
set $\alpha_k = 1$, $k = 1, \ldots, K$, so that the covariance-matrix estimator
\eqref{eq:hdrda-cov} resembled \eqref{eq:ridge-estimator}. We estimated
$\lambda$ from a grid of 21 equidistant candidate values between 0 and 1,
inclusively. Similarly, we estimated $\gamma$ from a grid consisting of the
values $10^{-1}, \ldots, 10^4$, and $10^5$. We selected optimal estimates of
$\lambda$ and $\gamma$ using $V$-fold cross-validation with $V = 10$. For the
second model, we set $\alpha_k = 1 - \gamma$, $k = 1, \ldots, K$, to resemble
\citeauthor{Friedman:1989tm}'s parameterization, and we estimated both $\lambda$
and $\gamma$ from a grid of 21 equidistant candidate values between 0 and 1,
inclusively.

We did not include the \emph{RDA} classifier in our classification study because
its training runtime was prohobitively slow on high-dimensional data in our
preliminary experiments. As shown in Section \ref{sec:timing-comparisons}, the
runtime of the \emph{RDA} classifier was drastically larger than that of the
\emph{HDRDA} classifier for a tuning grid of size $25 = 5 \times
5$. Consequently, a fair comparison between the \emph{RDA} and \emph{HDRDA}
classifiers would require model selection of $441 = 21 \times 21$ different
pairs of tuning parameters in the \emph{RDA} classifier. A tuning grid of this
size yielded excessively slow training runtimes for the \emph{RDA}
implementation from the {\tt klaR} R package.
